---
title: "Using grEPI to Create Epidemiological Parameter Summary Tables"
author: "Steve Kerr"
date: 11 June 2025
output: 
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  message = FALSE, 
  warning = FALSE
)
```

## Overview

### What is grEPI?

grEPI (Global Repository of Epidemiological Parameters) is WHO’s centralized database containing systematically reviewed epidemiological parameters for various pathogens. This tutorial shows how to access and summarize these parameters programmatically.

### Learning Objectives

- Connect to WHO Collaboratory grEPI API endpoints
- Extract and clean epidemiological parameters from API responses
- Create summary tables for pathogen parameters

### Prerequisites

- Basic R knowledge
- Understanding of epidemiological concepts


### Required Libraries

```{r, message=FALSE}
library(httr)       # For HTTP requests to APIs
library(jsonlite)   # To parse JSON responses from APIs
library(tidyverse)  # For data wrangling and visualization
library(janitor)    # For cleaning column names and data inspection
library(gt)         # For nicely formatted table output
```


## 1. Explore Available Endpoints

We begin by querying the root of the WHO Collaboratory API to discover which datasets are available. This helps us identify the relevant table containing mpox parameters.

```{r}
url <- "https://xmart-api-public-uat.who.int/COLLAB"
response <- GET(url)

if (status_code(response) != 200) {
  stop("API request failed with status: ", status_code(response))
}

json_text <- content(response, "text")
data <- fromJSON(json_text)
endpoint_df <- data$value
```

```{r, echo=FALSE}
gt(endpoint_df) %>% opt_interactive(use_search = TRUE, use_filters = TRUE, page_size_default = 10)
```

<div class="alert alert-info" role="alert">

💡 Look for datasets ending in `_PARAMETERS` for epidemiological data

</div>

## 2. Extract Mpox Parameters

Once we have the endpoint name, we connect specifically to the `MPOX_PARAMETERS` table. We then clean the raw data by removing system-generated columns.

```{r}
url <- "https://xmart-api-public-uat.who.int/COLLAB/MPOX_PARAMETERS"
response <- GET(url)

if (status_code(response) != 200) {
  stop("Failed to fetch mpox data. Status: ", status_code(response))
}

json_text <- content(response, "text")
json_data <- fromJSON(json_text)
mpox_parameters_table <- json_data$value

mpox_raw <- mpox_parameters_table %>%
  clean_names() %>%
  select(-starts_with("sys_"), -record_id)
```

✓ Raw data extracted: `r nrow(mpox_raw)` records

```{r, echo=FALSE}
gt(mpox_raw) %>% opt_interactive(use_search = TRUE, use_filters = TRUE, page_size_default = 3)
```


We confirm that the data was successfully extracted and then focus only on fields relevant for analysis: parameters, population characteristics, and source articles.

```{r}
glimpse(mpox_raw)

mpox_df <- mpox_raw %>%
  select(
    starts_with("param_"),
    distribution_type_fk,
    starts_with("pop_"),
    starts_with("article_")
  )
```

✓ Focused dataset created with `r ncol(mpox_df)` relevant columns

```{r, echo=FALSE}
gt(mpox_df) %>% opt_interactive(use_search = TRUE, use_filters = TRUE, page_size_default = 3)
```


## 3. Filter for Incubation Period Parameters

Next, we filter the dataset to retain only records related to the incubation period. We restrict results to human delay parameters measured in days.

```{r}
tabyl(mpox_df$param_subtype_fk)

mpox_ip <- mpox_df %>%
  filter(
    param_type_fk == "Human delay",
    param_subtype_fk %in% c("incubation period", "incubation period (infection to onset)"),
    param_unit_fk == "Days"
  )
```

✓ Filtered to `r nrow(mpox_ip)` incubation period records

```{r, echo=FALSE}
gt(mpox_ip) %>% opt_interactive(use_search = TRUE, use_filters = TRUE, page_size_default = 3)
```


`r if (nrow(mpox_ip) == 0) warning("No incubation period data found. Check filter criteria.")`

## 4. Data Cleaning and Transformation

This step formats the data for analysis and visualization. We:
	- Convert numeric columns to the right type
	- Round values
	- Create readable age and uncertainty ranges
	- Harmonize uncertainty types (e.g., “CRI95%” becomes “95% CrI”)

```{r}
mpox_clean <- mpox_ip %>%
  mutate(
    across(matches("^param_uncert_.*_val$"), as.numeric),
    across(matches("^pop_age_"), as.integer),
    across(where(is.numeric), ~ round(.x, 1)),
    param_value_type_fk = str_to_title(param_value_type_fk),
    range = case_when(
      !is.na(param_bound_lower) & !is.na(param_bound_upper) ~ paste(param_bound_lower, "-", param_bound_upper),
      TRUE ~ NA_character_
    ),
    uncert_bounds = case_when(
      !is.na(param_uncert_lower_val) & !is.na(param_uncert_upper_val) ~ paste(param_uncert_lower_val, "-", param_uncert_upper_val),
      TRUE ~ NA_character_
    ),
    pop_age_range = case_when(
      !is.na(pop_age_min) & !is.na(pop_age_max) ~ paste(pop_age_min, "-", pop_age_max),
      !is.na(pop_age_min) ~ paste0(pop_age_min, "+"),
      TRUE ~ NA_character_
    ),
    param_uncert_type_fk = case_when(
      param_uncert_type_fk == "CRI95%" ~ "95% CrI",
      param_uncert_type_fk == "CI95%" ~ "95% CI",
      TRUE ~ param_uncert_type_fk
    )
  )
```

```{r, echo=FALSE}
gt(mpox_clean) %>% opt_interactive(use_search = TRUE, use_filters = TRUE, page_size_default = 3)
```

```{r}
mpox_pivot <- mpox_clean %>%
  pivot_wider(
    names_from = param_uncert_type_fk,
    values_from = uncert_bounds
  )
```

```{r, echo=FALSE}
gt(mpox_pivot) %>% opt_interactive(use_search = TRUE, use_filters = TRUE, page_size_default = 3)
```

We then reshape the data into a wide format so that each uncertainty type (e.g., IQR, 95% CI) gets its own column.


## 5. Create Final Summary Table

We prepare a tidy summary table with selected variables that are ready for interpretation or export. This includes the parameter value, uncertainty bounds, distribution, and population characteristics.

The table below is interactive — you can search, sort, and scroll horizontally.

```{r}
result <- mpox_pivot %>%
  select(
    `Article Label` = article_label,
    `Article Title` = article_title,
    `Country` = pop_country,
    `Sample Size` = pop_sample_size,
    `Sex` = pop_sex_fk,
    `Age Range` = pop_age_range,
    `Parameter Type` = param_value_type_fk,
    `Value (days)` = param_val,
    `95% CrI`,
    `95% CI`,
    `IQR`,
    `Range` = range,
    `Distribution` = distribution_type_fk
  ) %>%
  arrange(`Article Label`) %>%
  select(where(~ !all(is.na(.))))

# Optional: Save results
# write_csv(result, "mpox_incubation_parameters.csv")


result %>%
  gt() %>%
  tab_header(
    title = md("**Mpox Incubation Period Parameters from grEPI**"),
  ) %>%
  fmt_missing(
    columns = everything(),
    missing_text = "—"
  ) %>%
  opt_table_outline() %>%
  opt_interactive(use_search = TRUE, use_filters = TRUE, page_size_default = 5)
```


Summary of incubation period parameters found in the table:

- **Total studies included:** `r nrow(result)`
- **Countries represented:** `r length(unique(result$Country[!is.na(result$Country)]))`
- **Mean incubation period:** `r round(mean(result[["Value (days)"]], na.rm = TRUE), 1)` days
- **Range of values:** `r round(min(result[["Value (days)"]], na.rm = TRUE), 1)` – `r round(max(result[["Value (days)"]], na.rm = TRUE), 1)` days
 

<div class="alert alert-warning" role="alert">

**Troubleshooting Tips:**

- If API fails: Check internet connection and try again  
- If no data returned: Verify filter criteria match available values  
- For other pathogens: Replace `'MPOX_PARAMETERS'` with desired endpoint

</div>

<div class="alert alert-info" role="alert">

**Try These Exercises:**

- Modify the script to extract serial interval parameters instead  
- Filter for a specific country or age group

</div>
